---
title: "Reproducible Research : Peer Assessment 1"
output:
    html_document:
        keep_md: true
---

# Course Project - I (Reproducible Research)

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Load and Pre-processing the data

```{r loading data}
library(dplyr)
library(ggplot2)

activity <- read.csv("activity.csv")

activity$date <- as.Date(activity$date)

str(activity)
```


### What is mean total number of steps taken per day?

```{r steps}
daily_steps <- activity %>%
    group_by(date) %>%
    summarise(total_steps = sum(steps, na.rm = T))

ggplot(daily_steps, aes(x = total_steps)) +
    geom_histogram(binwidth = 1000, fill = "blue", color = "black") +
    labs(title = "Histogram of Total steps taken each day", 
         x = "Total Steps per Day",
         y = "Frequency")

mean_steps <- mean(daily_steps$total_steps, na.rm = T)
median_steps <- median(daily_steps$total_steps, na.rm = T)
```

The mean number of steps taken per day is `r mean_steps` and the median is `r median_steps`.


### What is the averge daily activity pattern ?

```{r daily pattern}
interval_avg <- activity %>% 
    group_by(interval) %>%
    summarise(avg_steps = mean(steps, na.rm = T))

ggplot(interval_avg, aes(x = interval, y = avg_steps)) + 
    geom_line(color = "blue") + 
    labs( title = "Average Daily activity pattern", 
          x = "5-minute Interval",
          y = "Average Number of Steps")

max_interval <- interval_avg %>% filter(avg_steps == max(avg_steps))
```

The 5-minute interval with the maximum average number of steps is `r max_interval$interval` with `r round(max_interval$avg_steps,2)` steps on average.


### Imputing Missing Values

```{r missing values}
total_na <- sum(is.na(activity$steps))

activity_imputed <- activity %>%
    group_by(interval) %>%
    mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = T), steps)) %>%
    ungroup()

daily_steps_imputed <- activity_imputed %>%
    group_by(date) %>%
    summarise(total_steps = sum(steps))

ggplot(daily_steps_imputed, aes(x = total_steps)) +
    geom_histogram(binwidth = 1000, fill = "red", color = "black") +
    labs(title = "Histogram of Total steps taken each day (Imputed Data)", 
         x = "Total steps per Day",
         y = "Frequency")

mean_steps_imputed <- mean(daily_steps_imputed$total_steps)
median_steps_imputed <- median(daily_steps_imputed$total_steps)
```

There are `r total_na` missing values in the dataset.

After imputing missing data with the mean for each 5-minute interval:

- The new mean is `r mean_steps_imputed` (originally the mean was `r mean_steps`)
- The new median is `r median_steps_imputed` (originally the median was `r median_steps`)

The impact of imputing missing data was to increase both the mean and median, as the original method treated days with all NA values as having 0 steps, which pulled down the averages.


### Are there differences in activity patterns between weekdays and weekends ?

```{r differences in weekdays and weekends}
activity_imputed <- activity_imputed %>%
    mutate(day_type = ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "weekend", "weekday")) %>%
    mutate(day_type = as.factor(day_type))

interval_avg_daytype <- activity_imputed %>%
    group_by(interval, day_type) %>%
    summarise(avg_steps = mean(steps))

ggplot(interval_avg_daytype, aes(x = interval, y = avg_steps)) +
    geom_line(color = "blue") +
    facet_grid(day_type ~ .) +
    labs(title = "Average Daily Activity Pattern: Weekday vs. Weekend", 
         x = "5-minute Interval",
         y = "Average Number of Steps")
```

The plot shows clear differences between weekdays and weekends. Weekdays show a pronounced morning peak, while weekends show more consistent activity throughout the day.
